\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-09-23 chapters/bci.tex
Comments Liesel

\todo{
Locked-in -\> communicate
Paraplegic -\> walk again
tetraplegic, locked-in syndrome
Regaining capabilities and improving quality of life
Start from wolpaw examples, for spelling
Cite some catchy state of the art examples (walk again, speech avatar Metzger, handwriting
willet, neuralink)
}
\todo{put focus on communication BCI}
\section{A direct interface to the brain}
\comment{This section describes a bci in layman's terms, without too much
scientific background, jargon and literature, in the following sections of this
chapter a more scientific style is used}
A \ac{bci} reads your mind.
\ac{bci}s are an entertaining subject with many futuristic and more established
applications.
Devices processing direct inputs from the central nervous system can be
useful for rehabilitation, medical diagnosis or treatment.
They have always been especially promising as assistive technology for
paralyzed patients limited in their communication ability. \todo{expand more on this last
sentence, put patients first in introduction, see above}
Furthermore, they bring a new paradigm to human-computer interaction,
especially when paired with virtual or augmented reality.

@@

Let's break this definition down and focus on its separate parts.
First of all, we need to identify a signal that is a direct representation of
what's going on in the brain.
Section~\ref{sec:bci-recording} gives an overview of the available options.
One might for instance measure the fluctuations in bloodflow to specific brain
regions that are more or less active at a given time with a scanner.
However, this signal reacts to slowly to reflect the real-time activity that is
of interest for our purposes, and it carries little information other than the
brain regions where it is originating.
A better candidate is the neuronal \emph{action potential}.
The brain consists 86 billion neurons, which are highly interconnected cells that are
the smallest units of information processing.
The action potential is an electrical pulse occurring when a neuron receives
input from another neuron and is activated.\todo{gross oversimplification,not
entirely true}
A neuron's action potential, or the combined action potentials of groups of
synchronized neurons, thus generate an electrical field in and near the brain,
which can be measured using electrodes.
Fluctuations in the brain's electrical potential can be faster and more
informative.

@@

The previous points to another component of a successful \ac{bci}: the specific
brain signals related to our paradigm need to be identified within the recorded
activity.
\todo{preprocessing}
This is commonly referred to as \emph{feature extraction}
In other words, the signal needs to be interpreted \emph{in function of} the
task we want our \ac{bci} to perform.
The electrical activity of the brain is relatively weak compared to that of the
environment around us, containing electronic devices and other sources of
interference.
Furthermore, the brain is continuously processing information and carrying out
`background tasks', all of which generate their own electrical signals.
On the other hand, the brain activity we are interested in exploiting often
originates from a specific functional system within the brain, and its activity
might not be easily discerned from all other neural and environmental activity,
akin to trying to listen to a single speaker in a noise environment where
everyone is shouting over each other.
Some obvious sources of interference can be easily filtered out, but to obtain
a good measure reflecting only the activity that is relevant for a given \ac{bci}
problem.
This problem is called \emph{decoding} and is often solved through supervised
machine learning.
Useful information that is encoded in the brain signal by the paradigm,
can be retrieved from it by the decoder.
Section~\label{sec:bci-decoding} takes a closer look at some common decoding
techniques.

Finally, the loop can be closed by coupling this to a
device or actuator.
There are two aspects to this.
On the one hand, the \ac{bci} gains its function by allowing the user to interact
with their environment.\todo{put focus on communication BCI}
On the other hand, the actions performed by the \ac{bci} can themselves influence
the user's brain activity, creating an adaptive system.
A direct form of closed-loop \ac{bci} is the use of a neurostimulator.
Examples of this are deep brain stimulation to mitigate the symptoms of
Parkinson's disease, or the cochlear implant as an auditive prosthesis.
More indirectly modulation can be achieved by sensory stimulation.
This can involve specific, paradigm-related sensory input, like haptic feedback when a
movement \ac{bci} detects an intended action, or simply presenting selected actions
back to the user.
The user's brain will then adapt through reinforcement learning, causing
changes in behavior or strategy.
Due to neuroplasticity, the brain's ability to adapt, this can have a positive
impact on \ac{bci} performance and opens the pathway to rehabilitation applications.

@@

The brain's activity can be recorded using various neuroimaging
technologies.
These can range from brain scans~\cite{Weiskopf2004} using functional magnetic resonance imaging
(fMRI) to more portable technologies like accoustic signals obtained by functional ultrasound imaging
(fUS)~\cite{Zheng2023} and near-infrared spectroscopy
(fNIRS)~\cite{Borgheai2020}.
These technologies all measure the hemodynamic signal, which only indirectly
reflects brain activity.
Because of the desirable high temporal resolution of
electrophysiology~\cite{Easttom2021} and
because of their practicality, we will focus only on methods to record
electromagnetical activity originating from neuronal action potentials.
The magnetical field of the brain can be measured using magnetoencephalography
(MEG)~\cite{Mellinger2007}.
While MEG is non-invasive and results in high-quality signals, it is relatively
impractical due to the size of the equipment needed.
Recent advances have been made using optically pumped magnetometers
(OPM-MEG)~\cite{Wittevrongel2021}, but this technology still falters outside of
lab conditions.

@@

Larger implanted electrodes are referred to as intracranial ac{eeg} (iac{eeg}).
Depth electrodes such as those used in stereo-ac{eeg} \ac{bci}~\cite{Wu2024} tasks
and closed-loop adaptive control for Deep Brain Stimulation to mitigate
Parkinson's disease symptoms~\cite{Arlotti2018}.
Since these still penetrate the cortex, electrodes that sit on top of the
cortex are more popular, as they are considered less invasive.
\ac{ecog} is a powerful method
with many applications in \ac{bci}~\cite{Schalk2011} and epilepsy diagnosis.
Newer developments like micro-ECoG (µECoG)~\cite{Shokoueinejad2019} with a
higher number of smaller electrodes per surface area allow for more precise
measurements.

@@

\Ac{eeg} is the de facto non-invasive electrophysiology standard, in clinical
neurology practice as well as in \ac{bci} research.
Developed in 1924~\cite{Berger1929}, it is cheap and relatively practical and
can offer great temporal resolution up to thousands of Hz.
As it is easily applicable on the outside of the scalp, many practical
consumer grade \ac{eeg} headsets exist in the form (often wireless) electrode
caps or helmets.
Current consumer systems can feature dry electrodes, but clinical and
research systems often use electrolyte gel to reduce impedance for improved signal
quality, which slightly decreases practicality.
Nevertheless, it is the most accessible \ac{bci} technology for both patients
and researchers.

The major drawback of \ac{eeg} is its poor spatial resolution.
The electrodes are large and sit on the outside of the scalp, relatively far
away from the cortex.
Furthermore, the layers of skull and cerebrospinal fluid in between can
attenuate the signal and cause volume conduction, which results in electrodes
measuring a mixture of activity from sources elsewhere in the brain.
In addition to picking up brain activity other than that of the region of
interest, \ac{eeg} also records other artifacts, from the
environment, the power supply, nearby electrical equipment or the user's muscle
movements.
Combined, this means that \ac{eeg} is inherently noisy.

Despite its noisy nature \Ac{eeg} is the recording methodology of choice for
our \ac{bci} because of its wide acceptance by patients.
To counteract the noise present in our recording, we must on the one hand evoke
strong, informative brain signals using a suited \ac{bci} \emph{paradigm}, and on th
other hand pick a suited \emph{decoder} which can isolate the signal of
interest and filter out the noise.

\section{\ac{bci} Paradigms}
\label{sec:bci-paradigms}

@@

\section{Decoders}
\label{sec:bci-decoders}

After extracting features, machine learning algorithms can perform \ac{bci}
decoding.
This is usually based on
some training data for which the performed tasks are known, by recognize the
activity evoked by the task in recorded data.
The training data can be either obtained from the \ac{bci} user themselves or from
other users.
In the first case, the user would perform a short calibration session before
using the \ac{bci}, where they are instructed to perform a known task.
This calibration session can be eliminated by training the decoder on
preexisting labeled data from other sessions and users, but this is harder due
to the variability between subjects and sessions.\todo{make shorter, move part
to section}

@@

\textcite{Lotte2018, Xu2021} present a relatively recent overview of state-of-the art
decoder algorithms.
Classic linear methods, such as \ac{csp} feature extraction for
motor imagery~\cite{Park2017} and \ac{cca} for \ac{ssvep}~\cite{Nakanishi2017},
and \ac{lda} for \ac{erp} classification~\cite{Sosulski2022} still perform
relatively well, given some regularizing constraints or extensions.
Multi-linear techniques exploiting the tensor structure of neural signals are
also promising.
Riemannian Geometry~\cite{Barachant2014} is a popular and robust new strategy,
Furthermore, they lend themselves well to application like adaptive
learning~\cite{Benaroch2021} or transfer learning~\cite{Zanini2017}
Riemannian Geometry classifiers are often considered the current
state-of-the-art.
Finally, deep learning\todo{Bhuvaneshwari2021} is also sometimes considered, albeit only when large
training datasets are available.
If decoders tailored to a specific user that keep the calibration session as
short as possible are of interest, not enough training data is available to
properly train a deep learning model.

Usually, a decoder makes a prediction for a given \emph{trial} while operating
the \ac{bci}.
A trial is the smallest unit on which a selection decision can be made, for
instance be one imagined movement, or one repetition of flashing all targets
in a visual \ac{erp} \ac{bci}.
Some valuable metrics to asses decoder performance in the classification case
are accuracy and \ac{rocauc}.
Accuracy can be obtained from the correctness of predicted trial selection as
\begin{equation}
  \text{accuracy} = \frac{\text{TP}+\text{TN}}{\text{TP}+\text{TN}+\text{FP}+\text{FN}}
\end{equation}
with TP and TN the numbers of true positives and negatives and FP and FP the
numbers of false positives and negatives.
\Ac{rocauc} is also a measure of classifier predictive power, but is more
suited for evaluation of classification of single epochs of data and in the
presence of imbalanced data, e.g. when
comparing single \acp{erp}.
Higer \ac{rocauc} usually translates to a higher target selection accuracy.

>>>>

\todo{
Locked-in -\> communicate
Paraplegic -\> walk again
tetraplegic, locked-in syndrome
Regaining capabilities and improving quality of life
Start from wolpaw examples, for spelling
Cite some catchy state of the art examples (walk again, speech avatar Metzger, handwriting
willet, neuralink)
}
\todo{put focus on communication BCI}
\section{A direct interface to the brain}
\comment{This section describes a bci in layman's terms, without too much
scientific background, jargon and literature, in the following sections of this
chapter a more scientific style is used}
A \ac{bci} reads your mind.
\ac{bci}s are an entertaining subject with many futuristic and more established
applications.
Devices processing direct inputs from the central nervous system can be
useful for rehabilitation and medical diagnosis or treatment.
They have always been especially promising as assistive technology for
paralyzed patients limited in their communication ability. \todo{expand more on this last
sentence, put patients first in introduction, see above}
Furthermore, they bring a new paradigm to human-computer interaction,
especially when paired with virtual or augmented reality.

@@

Let's break this definition down and focus on its separate parts.
First of all, we need to identify a signal that is a direct representation of
what's going on in the brain.
Section~\ref{sec:bci-recording} gives an overview of the available options.
One might for instance measure the fluctuations in bloodflow to specific brain
regions that are more or less active at a given time with a scanner.
However, this signal reacts too slowly to reflect the real-time activity that is
of interest for our purposes, and it carries little information other than the
brain regions where it is originating.
A better candidate is the neuronal \emph{action potential}.
The brain consists 86 billion neurons, which are highly interconnected cells that are
the smallest units of information processing.
The action potential is an electrical pulse occurring when a neuron receives
input from another neuron and is activated.\todo{gross oversimplification,not
entirely true}
A neuron's action potential, or the combined action potentials of groups of
synchronized neurons, thus generate an electrical field in and near the brain,
which can be measured using electrodes.
Fluctuations in the brain's electrical potential can be faster and more
informative.

@@

The previous points to another component of a successful \ac{bci}: the specific
brain signals related to our paradigm need to be identified within the recorded
activity.
\todo{preprocessing}
This is commonly referred to as \emph{feature extraction}
In other words, the signal needs to be interpreted \emph{in function of} the
task we want our \ac{bci} to perform.
The electrical activity of the brain is relatively weak compared to that of the
environment around us, containing electronic devices and other sources of
interference.
Furthermore, the brain is continuously processing information and carrying out
`background tasks', all of which generate their own electrical signals.
On the other hand, the activity of interest in often
originates from a specific functional system within the brain.
This activity might not easily be discerned from all other neural and
environmental activity.
Compare this to trying to listen to a single speaker at a noisy party where
everyone is shouting over each other.
Some obvious sources of interference can be easily filtered out by
\emph{preprocessing}, but the problem of retrieving or \emph{decoding} only the
relevant signal is often solved through supervised machine learning.
Useful information that is encoded in the brain signal by the paradigm,
can be retrieved from it by the decoder.
Section~\label{sec:bci-decoding} takes a closer look at some common decoding
techniques.

Finally, the loop can be closed by coupling this to a
device or actuator.
There are two aspects to this.
On the one hand, the \ac{bci} gains its function by allowing the user to interact
with their environment.\todo{put focus on communication BCI}
On the other hand, the actions performed by the \ac{bci} can themselves alter
the user's cognitive state or brain activity, creating an adaptive system.
A direct form of closed-loop \ac{bci} is the use of a neurostimulator.
Examples of this are deep brain stimulation to mitigate the symptoms of
Parkinson's disease, or the cochlear implant as a hearing prosthesis.
More indirectly, modulation can be achieved by sensory stimulation.
This involves specific, paradigm-related sensory input, like haptic selection
feedback for movement \ac{bci}, or simply presenting selected actions
back to the user.
The user's brain will then adapt through reinforcement learning, causing
changes in behavior or strategy.
Neuroplasticity, the brain's ability to adapt and reinforce specific neural
pathways, can have a positive impact on \ac{bci} performance and gives rise to
rehabilitation applications.

@@

The brain's activity can be recorded using various neuroimaging
technologies.
These can range from brain scans~\cite{Weiskopf2004} using functional magnetic resonance imaging
(fMRI) to more portable technologies like accoustic signals obtained by functional ultrasound imaging
(fUS)~\cite{Zheng2023} and near-infrared spectroscopy
(fNIRS)~\cite{Borgheai2020}.
The technologies above all measure the hemodynamic signal, which only indirectly
reflects brain activity.
Because of the desirable high temporal resolution~\cite{Easttom2021} and
practicality electrophysiology, we will focus only on methods to record
electromagnetical activity originating from neuronal action potentials.
The magnetical field of the brain can be measured using magnetoencephalography
(MEG)~\cite{Mellinger2007}.
While MEG is non-invasive and results in high-quality signals, it is relatively
impractical due to the size of the equipment needed.
Recent advances have been made using optically pumped magnetometers
(OPM-MEG)~\cite{Wittevrongel2021}, but this technology still falters outside of
lab conditions.

@@

Larger implanted electrodes are referred to as intracranial ac{eeg} (iac{eeg}).
Depth electrodes such as those used in stereo-ac{eeg} \ac{bci}~\cite{Wu2024} tasks
and closed-loop adaptive control for Deep Brain Stimulation to mitigate
Parkinson's disease symptoms~\cite{Arlotti2018}.
Since these still penetrate the cortex, electrodes that sit on top of the
cortex are more popular, as they are considered less invasive.
\Ac{ecog} is a powerful method
with many applications in \ac{bci}~\cite{Schalk2011} and epilepsy diagnosis.
Newer developments like micro-ECoG (µECoG)~\cite{Shokoueinejad2019} with a
higher number of smaller electrodes per surface area allow for more precise
measurements.

@@

\Ac{eeg} is the de facto non-invasive electrophysiology standard in clinical
neurology practice as well as in \ac{bci} research.
Developed since 1924~\cite{Berger1929}, it is cheap and relatively practical and
offer great temporal resolution up to thousands of Hz.
As it is easily applicable on the outside of the scalp, many practical
consumer-grade \ac{eeg} headsets exist in the form (often wireless) electrode
caps or helmets.
Current consumer systems sometimes feature dry electrodes, but clinical and
research systems often use electrolyte gel to reduce impedance for improved signal
quality, which slightly decreases practicality.
Nevertheless, it is the most accessible \ac{bci} technology for both patients
and researchers.

The major drawback of \ac{eeg} is its poor spatial resolution.
The electrodes are large and sit on the outside of the scalp, relatively far
away from the cortex.
Furthermore, the layers of skull and cerebrospinal fluid in between can
attenuate the signal and cause volume conduction, which results in electrodes
measuring a mixture of activity from sources elsewhere in the brain.
The noise problem mentioned earlier is also prominently present in \ac{eeg}
recordings.
In addition to picking up brain activity other than that of the region of
interest, \ac{eeg} also records other artifacts, from the environment, the power
supply, nearby electrical equipment, or the user's muscle movements.
Combined, this means \ac{eeg} is inherently noisy.

Despite its noisy nature, \ac{eeg} is the recording methodology of choice for
our \ac{bci} because of its wide acceptance by patients.
To counteract the noise present in our recording, we must on the one hand evoke
strong, informative brain signals using a suited \ac{bci} paradigm, and on the
other hand pick a suited decoder which can isolate the signal of interest and
filter out the noise.

\section{\ac{bci} paradigms}
\label{sec:bci-paradigms}

@@

\section{Decoders}
\label{sec:bci-decoding}

After preprocessing and extracting features, machine learning algorithms can perform \ac{bci}
decoding.
This is usually based on
some training data for which the performed tasks are known, by recognizing the
activity evoked by the task in recorded data.
The training data can be either obtained from the \ac{bci} user themselves or from
other users.
In the first case, the user would perform a short calibration session before
using the \ac{bci}, where they are instructed to perform a known task.
This calibration session can be eliminated by training the decoder on
preexisting labeled data from other sessions and users, but this is harder due
to the variability between subjects and sessions.\todo{make shorter, move part
to section}

@@

\textcite{Lotte2018, Xu2021} present a relatively recent overview of state-of-the art
decoder algorithms.
Classic linear methods, such as \ac{csp} feature extraction for
motor imagery~\cite{Park2017} and \ac{cca} for \ac{ssvep}~\cite{Nakanishi2017},
and \ac{lda} for \ac{erp} classification~\cite{Sosulski2022} still perform
relatively well, given some regularizing constraints or extensions.
Multi-linear techniques exploiting the tensor structure of neural signals are
also promising.
Riemannian Geometry~\cite{Barachant2014} is a popular and robust new strategy.
Furthermore, they lend themselves well to application like adaptive
learning~\cite{Benaroch2021} or transfer learning~\cite{Zanini2017}.
Riemannian Geometry classifiers are often considered the current
state-of-the-art.
Finally, deep learning~\cite{Bhuvaneshwari2021} is also sometimes considered,
albeit only when sufficiently large training datasets are available.
If decoders tailored to a specific user that keep the calibration session as
short as possible are of interest, not enough training data is available to
properly train a deep learning model.

Usually, a decoder makes a prediction for a given \emph{trial} while operating
the \ac{bci}.
A trial is the smallest unit on which a selection decision can be made, for
instance be one imagined movement, or one repetition of flashing all targets
in a visual \ac{erp} \ac{bci}.
Some valuable metrics to asses decoder performance in the classification case
are accuracy and \ac{rocauc}.
Accuracy is calculated as the proportion of correct selections to all
selections made.
It should be carefully interpreted in the presence of imbalanced data and always compared to the
random chance accuracy level in the presence of more than two possible
selections per trial.
\Ac{rocauc} is also a measure of classifier predictive power, but is more
suited for evaluation of classification of single epochs of data and in the
presence of imbalanced data, e.g. when comparing single \acp{erp}.
Higher \ac{rocauc} usually translates to a higher target selection accuracy.
